---
title: "Session 2 Solutions"
author: "Tim Riffe"
date: "2023-07-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Best practice life exectancy:

Make a scatterplot of life expectancy, with males and females in different panels, drawn as a background point cloud (light gray and semitransparent or similar). For each panel and each year, highlight the highest observed life expectancy using red points. Add a fitted line to the series of record high life expectancies by sex. Add another fitted line to the overall point cloud of life expectancy. Add a narrative title.

Here's how to get the highest life expectancy per year and sex, in order to get you started. 
```{r}
library(tidyverse)
gapreminder <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2023Module1/master/Data/gapreminder.csv", show_col_types = FALSE) 

best_practice <-
  gapreminder |> 
  filter(year < 2022) |> 
  select(country, year, e0f, e0m) |> 
  pivot_longer(e0f:e0m, names_to = "sex", values_to = "e0") |> 
  mutate(sex = if_else(sex == "e0f","Women","Men")) |> 
  group_by(year, sex) |> 
  filter(e0 == max(e0, na.rm = TRUE)) |> 
  ungroup() |> 
  arrange(sex, year)
```

Now let's make the background point cloud:
```{r, warning = FALSE, message = FALSE}
gapreminder |> 
  select(year, e0f, e0m) |> 
  pivot_longer(-1, names_to = "sex", values_to = "e0") |> 
  mutate(sex = if_else(sex == "e0f","Women","Men")) |> 
  # here begins the plot
  ggplot(aes(x = jitter(year), y = e0)) +
  # the background point cloud:
  geom_point(color = gray(.4), 
             alpha = .1, 
             size = .5) +
  geom_smooth(method = "lm", 
              linewidth = .5, 
              color = "#132294") +
  # declare the facet
  facet_wrap(~sex) +
  
  # the best practice part
  geom_point(data = best_practice,
             color = "red",
             size = .5) +
  geom_smooth(data = best_practice,
              method = "lm", 
              color = "#941347",
              linewidth = .5) +
  theme_minimal() +
  labs(x = "year", 
       y = "Life Expectancy at Birth",
       title = "The global average life expectancy trend\nincreases faster than best practice life expectancy")
```
















