---
title: "Session 2 Notes"
author: "Tim Riffe"
date: "2023-07-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data wrangling

Today we'll see some tools that we can use to 
1. Read in spreadsheet data and csv data
2. select columns and filter rows
3. create or modify columns
4. reshape data (wide to long and vice versa)
5. recode data
And other tricks.

Finally, we'll merge all the datasets we manage to harmonize.

### WPP data
```{r}
library(tidyverse)
library(readxl)
```

For this spreadsheet, it's best to visually inspect to get a sense of what rows to skip and how to treat NAs, and possibly other oddities. Here we skip the first 16 rows, and declare that `"..."` should be treated as NA (as will empty cells).
```{r}
wpp <- read_excel("Data/WPP2022_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT_REV1.xlsx",
           skip = 16,
           na = "...")

nrow(wpp)
```

Next let's filter down to only those rows that have valid ISO3 codes.

```{r}
colnames(wpp)

wpp <-
  filter(wpp,
         !is.na(`ISO3 Alpha-code`))
nrow(wpp)
```

The function `is.na()` asks a vector which elements are NAs, returning TRUE for NAs and FALSE for everything else.

The operator `!` negates this, meaning that TRUE becomes FALSE and vice versa.

`filter()` selects rows using TRUE and FALSE.

Now we should select the columns we actually want:

```{r}
colnames(wpp)
wpp <-
  select(wpp,
         iso3 = `ISO3 Alpha-code`,
         year = Year,
         popm = `Male Population, as of 1 July (thousands)`,
         popf = `Female Population, as of 1 July (thousands)`,
         tfr = `Total Fertility Rate (live births per woman)`,
         e0m = `Male Life Expectancy at Birth (years)`,
         e0f = `Female Life Expectancy at Birth (years)`)
```
Note, we select columns and rename them at the same time, anything not named gets thrown out. We need to use back-ticks around arduous names that have spaces.

The point was made that overwriting a `wpp` object is risky and clunky. Some folks would do so in sequence, creating several unique staged versions of `wpp`, e.g. `wpp1`, `wpp2` ... Instead, we can string these statements, or steps into sentences. Or pipelines. Or single sequenced execution steps.
```{r}
wpp <- read_excel("Data/WPP2022_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT_REV1.xlsx",
           skip = 16,
           na = "...") |> 
  filter(!is.na(`ISO3 Alpha-code`)) |> 
  select(iso3 = `ISO3 Alpha-code`,
         year = Year,
         popm = `Male Population, as of 1 July (thousands)`,
         popf = `Female Population, as of 1 July (thousands)`,
         tfr = `Total Fertility Rate (live births per woman)`,
         e0m = `Male Life Expectancy at Birth (years)`,
         e0f = `Female Life Expectancy at Birth (years)`)
```

```{r}
ggplot(wpp, 
       mapping = aes(x = e0f,
                     y = e0m)) + 
  geom_point(mapping = aes(group = iso3)) + 
  geom_smooth(method = "lm")

wpp |> 
  ggplot(aes(x = e0f, y = tfr)) +
  geom_point(alpha = .2)

```


### GPI data

The GPI data we want are on the second sheet, so we specify `sheet = 2`, and to `skip` the first 3 rows. Otherwise, it's a simple format.

Then we need to convert the data rectangle to a tidy format, so we can make year a variable rather than having a unique column per year. This we do with the `pivot_longer()` function, which needs a column range (in this case skipping the first two columns), and a name to call the newly created *name collector* column (`names_to`), and place to store the values `values_to`. Then we follow with a select and rename step, and finally coerce `year` to integer using `as.integer()`. We do this inside of `mutate()`, which is used for modifying or creating columns.

```{r}
gpi <-
  read_excel("Data/GPI-2022-overall-scores-and-domains-2008-2022.xlsx",
             skip = 3,
             sheet = 2) |> 
  pivot_longer(-c(1,2), names_to = "year", values_to = "gpi") |> 
  select(iso3 = iso3c,
         year,
         gpi) |> 
  mutate(year = as.integer(year))
gpi
```

### World Bank data

We'll skip the selection of gender statistics downloaded from the World Bank (that was a haphazard selection from a very large set of possible variables). Instead let's read in the so-called `Popular Indicators` dataset.

Note: for the column range specification when pivotting longer, you can use names or column index numbers; You can also explicitly list them in `c()`, or using a range with `:`, and you can do so negatively (as below) or postively.
```{r}
wb <- 
  read_excel("Data/P_Popular Indicators.xlsx", 
             na = "..") |> 
  filter(`Series Code` %in% c("NY.GDP.PCAP.CD","EN.ATM.CO2E.PC")) |> 
  pivot_longer(-c(1:4), 
               names_to = "year", 
               values_to = "value") |> 
  select(-`Series Name`) |> 
  pivot_wider(names_from = `Series Code`, 
              values_from = value) |> 
  mutate(year = substr(year, start = 1, stop = 4),
         year = as.integer(year)) |> 
  select(iso3 = `Country Code`,
         year,
         co2pc = EN.ATM.CO2E.PC,
         gdppc = NY.GDP.PCAP.CD)
```

### Exercise: scatterplot
Make a scatterplot of GDPpc vs CO2pc, and see if you can also find a way to intuit the time trend in it using color or some other trick, like `alpha`.

New things in this plot:
We have a log log plot, which we do axis-by-axis. I chose a different plot theme (`theme_minimal()` is my go-to), and we gave explicit labels for the axes. In Monday's handout, I show how to format the gdp ticks labels better using the `scales` package.
```{r, warning = FALSE}
wb |> 
  ggplot(mapping = aes(
    x = gdppc,
    y = co2pc,
    color = year
  )) +
  geom_point(alpha = .3) +
  scale_y_log10() +
  scale_x_log10() + 
  theme_minimal() +
  labs(x = "GDP per capita",
       y = "CO2 Emissions per capita")
```

### ILO data

Here's a quick solution for the wage data.
```{r, warning = FALSE}
wage <- 
  read_csv("Data/SDG_0851_SEX_OCU_NB_A.csv.gz",
           show_col_type = FALSE) |> 
  filter(classif1 == "OCU_SKILL_TOTAL") |> 
  select(iso3 = ref_area,
         sex,
         year = time,
         value = obs_value) |> 
  pivot_wider(names_from = sex, 
              values_from = value) |> 
  mutate(wage_ratio = SEX_F / SEX_M) |> 
  select(iso3, year, wage_ratio)

```

Exercise: do this for the labor force data.

Note: the file starts with EAP_ ...
You'll want to filter down to the `classif1` called `MTS_AGGREGATE_TOTAL`.
Then you'll want to do something similar in order to have womens' labor force and mens' labor force sizes side by side in columns called `lff` and `lfm`.

The outgoing four variables will be `iso3`, `year`, `lff`, `lfm`.

In 5 minutes I'll derive a solution. Starting now.

```{r}
lf <-
  read_csv("Data/EAP_TEAP_SEX_MTS_NB_A.csv.gz",
         show_col_types = FALSE)  |> 
  filter(classif1 == "MTS_AGGREGATE_TOTAL") |> 
  select(iso3 = ref_area,
         year = time,
         sex,
         value = obs_value) |> 
  pivot_wider(names_from = sex,
              values_from = value) |> 
  select(iso3, year, lff = SEX_F, lfm = SEX_M)
```

## Merging

When joining (merging) datasets, you need to have variables to use as keys to match on. They should have conformable (matchable) values. In our case we needed to make sure year is always integer. And we were careful to always pick ISO3 codes.

In a left join, the dataset given first (on the left) never loses rows; new rows are also not generally created. All we do is augment the columns of the dataset on the left.

An inner join is the strictest kind of join, where only matches are kept (you might lose rows on the left and on the right).

A full join is lossless. Data can only stay the same size or get bigger.

```{r}
wpp <-
  wpp |> 
  mutate(year = as.integer(year))

left_join(gpi, wpp, by = c("iso3", "year"))
inner_join(gpi, wpp, by = c("iso3","year"))
full_join(gpi, wpp, by = c("iso3","year"))
```

```{r}
gapreminder <-
  wpp |> 
  full_join(gpi, by = c("iso3", "year")) |> 
  full_join(wb, by = c("iso3", "year")) |> 
  full_join(wage, by = c("iso3", "year")) |> 
  full_join(lf, by = c("iso3", "year"))
```

Exercise:
Is the Preston curve (with GDP logged) a stronger relation than CO2 per capita vs life expectancy? Should we use the total population life expectancy or female or male? Is the relationship stronger for females or males? Produce some visual evidence. I encourage the use of narrative plot titles.

```{r}
gapreminder |> 
  filter(year >= 2008) |> 
  select(e0f, co2pc, gdppc) |> 
  pivot_longer(2:3, names_to = "variable", values_to = "value") |> 
  ggplot(aes(x = value, y = e0f)) +
  geom_point(alpha = .1) +
  scale_x_log10() + 
  theme_minimal() +
  facet_wrap(~variable) +
  geom_smooth(method = "lm")
```















