---
title: "Session 3 Solutions"
author: "Tim Riffe"
date: "2023-07-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Review of functions:

A function has a 
1.    Name
2.    Arguments
3.    Body (what does it do with the arguments?)
4.    What it returns

You can make one yourself using this form:
```{r}
arbitrary_name <- function(arg1, arg2, ...){
  # body starts here
  # ....
  #
  return(result)
}
```



# Exercises

**\large Exercise 1.3.1: ** 
 
Read in the dataset `Data/hmd.csv.gz`:

```{r}
library(tidyverse)
hmd <- read_csv("https://github.com/timriffe/BSSD2023Module1/raw/master/Data/hmd.csv.gz")

```

This dataset is the entire HMD, where there is a lifetable subset for each unique combination of `country`, `sex`, and `year`. Complete the lifetable for the entire dataset **using your new function in a tidy pipeline**.

There are a couple aspects of syntax here, like the `~` and the name of the anonymous data chunk (of grouped data) passing through the pipeline (`.x`). I actually didn't remember these details when it came time to write this piece of code yesterday. To get that crucial detail, I mindlessly copied a piece of example code from the help file, modified to my data, and crossed my fingers. It worked. I got the help file like this: `?group_modify`

```{r}
library(tidyverse)
source("https://raw.githubusercontent.com/timriffe/BSSD2023Module1/master/my_functions.R")
hmd_completed <-
  hmd |> 
  group_by(country, sex, year) |> 
  group_modify(~ my_lifetable_tibble(data_chunk = .x)) |> 
  ungroup()
```

Let's have a look at life expectancy at birth:
```{r, warning = FALSE}

hmd_completed |> 
  filter(age == 0) |> 
  ggplot(aes(x = ex, fill = sex))+
  geom_density(alpha = .4)
```

**\large Exercise 1.3.2: ** 

$e(x)^\dagger$, pronounced e-dagger is the average year of life lost due to death in the lifetable, and one way to calculate it looks like this:

$$ e(x)^\dagger = \sum_{i=x}^\omega \frac{d(i)}{l(x)}  e(x) $$
That is to say, this is a function patterned by age $e(x)^\dagger$. If you want to calculate this for each age then you'll probably want to write a loop, which we've not done yet. But you can calculate it for age 0 without a loop, in which case (if the lifetable radix is 0):

 $$ e(0)^\dagger = \sum_0^\omega d(x)e(x)$$
Write a function that does this, and calculate $e(0)^\dagger$ for each Country, Sex, and Year. Also save $e(0)$, i.e. the value of $e(x)$ where `Age == 0`. Are you going to do this with `mutate()` or with `summarize()`? Can you plot the relationship between $e(0)$ and $e(0)^\dagger$? Make a scatterplot of this. If it's super overlapped try transparency. Maybe add a `geom_smooth()` on top of it, your choice of method.

```{r}
# this thing is not NA-proof. They'll mess it up.
calc_edag <- function(dx, ex){
  # remove radix, if it was used:
  dx    <- dx / sum(dx)
  
  # this is approximate
  edag0 <- sum(dx * ex)
  return(edag0)
}
```

`summarize()` is like `mutate()`, but for aggregation actions. Calculating a summary measure from a range of ages is a type of aggregation. It works like so:

```{r}
hmd_summary <-
  hmd_completed |> 
  group_by(country, sex, year) |> 
  summarize(e0 = ex[1],
            edag0 = calc_edag(dx,ex),
            .groups = "drop") 

```

Visualize it:

```{r, warning = FALSE}
hmd_summary |> 
  ggplot(mapping = 
           aes(x = e0, 
               y = edag0, 
               color = sex)) +
  geom_point(alpha = .1) +
  facet_wrap(~sex) +
  theme_minimal() +
  # how to turn of the legend created for a mapping
  guides(color = "none") +
  labs(title = "Usually improvement in life expectancy\nis paired with a decrease in lifespan inequality")
```


