---
title: "Session 3 Notes"
author: "Tim Riffe"
date: "2023-07-05"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions

What is a function? A function maps some inputs to an output. There
could be one or more inputs, and likewise for outputs.

-   `my_adder` is the name of the function
-   `function()` is the formal way to declare the named object as a
    function (rather than as data)
-   `()` we list the names of the arguments (parameters)
-   `{}` we create the function body; The function body only *sees* the
    arguments. It should not be written to use data objects sitting
    around in your R session. What happens in the function stays in the
    function. Don't be bashful. Main thing is it gets the job done.
-   `return()` is what comes out; nothing else comes out.

```{r}

my_adder <- function(number1, number2){
  the_sum <- number1 + number2
  return(the_sum)
}
my_adder(number1 = 3, number2 = 4)
my_adder(number1 = 3:10, number2 = 4:11)
```

Another function example: Let's make a Gompertz mortality function.
$$ \mu(x) = \alpha e^{\beta x}$$

-    $\mu(x)$ is the force of mortality at age $x$

-   $\alpha$ is a baseline scalar of mortality

-    $\beta$ is the pace of senescence (the proportional increase in
    mortality in each age step)

-   $x$ is age

```{r}
gomp_mx <- function(a, b, age){
  mx <- a * exp(b * age)
  return(mx)
}
```

These are our test parameters
```{r}
age     <- 0:110
alpha   <- 0.0001
beta    <- .08
library(tidyverse)
toymx <- 
  tibble(age = age,
         mx = gomp_mx(a = alpha,
                      b = beta, 
                      age = age))
```

## $q(x)$

$$ q(x) = \frac{\mu(x)}{1 + (1 - a(x)) \cdot \mu(x)}   $$
Make a function called `mx_to_qx()`. It will need arguments `mx` and `ax`. Go ahead

```{r}
ax <- rep(.5, times = 111)
mx <- toymx |> pull(mx)
# mx <- gomp_mx(a,b,age)
mx_to_qx <- function(mx, ax){
  qx <- mx / (1 + (1 - ax) * mx)
  qx[length(qx)] <- 1
  return(qx)
}
qx <- mx_to_qx(mx, ax)
```
Note: Why 111? Because the toy data goes up to age 110, and don't forget age 0 when counting elements.
`rep()` repeats a value `times` times.

`mx <- toymx |> pull(mx)` creates a vector called `mx` by pulling off a column of `toymx`. We could have also just made `mx` directly by using that original function not inside a tibble: `mx <- gomp_mx(a,b,age)`. I just opted to work with toy vectors rather than a tibble, sorry for that.

How to use a function:
```{r}
# toymx |> 
#   mutate(ax = .5,
#          qx = mx_to_qx(mx = mx, ax = ax))
qx <- mx_to_qx(mx, ax)
```

# Convert $q(x)$ to $l(x)$

$$ l(x) = \prod _{i=0}^x 1 - q(i) $$
Except the very first $l(x)$ should be a clean number like 1. 
I suggest using the `cumprod()`.
And recall that you can append a value to a vector with `c()`
Call the function `qx_to_lx()`

```{r}
qx_to_lx <- function(qx, radix = 1){
  lx <- cumprod(1 - qx)
  lx <- c(1,lx)
  lx <- lx[1:length(qx)]
  lx <- radix * lx
  return(lx)
}
lx <- qx_to_lx(qx)
# sum(lx)
# sum(lx) - .5
```

## $d(x)$ is the death distribution

$$d(x) = l(x)q(x)$$
It's also the same as:

$$ d(x) = l(x+1) - l(x)$$
$$ d(x) = -(l(x) - l(x+1)) $$
However, these last two are bit more tedious to code because you need to think about vector lengths. For the first one using `lx` an `qx` everything is already the right length.

```{r}
lxqx_to_dx <- function(lx,qx){
  lx * qx
}
dx <- lxqx_to_dx(lx,qx)
```

