---
title: "Session 3 Solutions"
author: "Tim Riffe"
date: "2023-07-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Exercises

**\large Exercise 1.3.1: ** 
 
Read in the dataset `Data/hmd.csv.gz`:

```{r}
library(tidyverse)
hmd <- read_csv("https://github.com/timriffe/BSSD2023Module1/raw/master/Data/hmd.csv.gz")
```

This dataset is the entire HMD, where there is a lifetable subset for each unique combination of `country`, `sex`, and `year`. Complete the lifetable for the entire dataset **using your new function in a tidy pipeline**.

```{r}
library(tidyverse)
source("https://raw.githubusercontent.com/timriffe/BSSD2023Module1/master/my_functions.R")
hmd_completed <-
  hmd |> 
  group_by(country, sex, year) |> 
  group_modify(~my_lifetable_tibble(.x)) |> 
  ungroup()
```

Let's have a look at life expectancy at birth:
```{r, warning = FALSE}
hmd_completed |> 
  filter(age == 0) |> 
  ungroup() |> 
  ggplot(aes(x=ex, fill = sex))+
  geom_density(alpha = .4)
```


**\large Exercise 1.3.2: ** 

$e(x)^\dagger$, pronounced e-dagger is the average year of life lost due to death in the lifetable, and one way to calculate it looks like this:

$$ e(x)^\dagger = \sum_{i=x}^\omega \frac{d(i)}{l(x)}  e(x) $$
That is to say, this is a function patterned by age $e(x)^\dagger$. If you want to calculate this for each age then you'll probably want to write a loop, which we've not done yet. But you can calculate it for age 0 without a loop, in which case (if the lifetable radix is 0):

 $$ e(0)^\dagger = \sum_0^\omega d(x)e(x)$$
Write a function that does this, and calculate $e(0)^\dagger$ for each Country, Sex, and Year. Also save $e(0)$, i.e. the value of $e(x)$ where `Age == 0`. Are you going to do this with `mutate()` or with `summarize()`? Can you plot the relationship between $e(0)$ and $e(0)^\dagger$? Make a scatterplot of this. If it's super overlapped try transparency. Maybe add a `geom_smooth()` on top of it, your choice of method.

```{r}
calc_edag <- function(dx,ex){
  # remove radix, if it was used:
  dx    <- dx / sum(dx)
  
  # this is approximate
  edag0 <- sum(dx * ex)
  return(edag0)
}
```

`summarize()` is like `mutate()`, but for aggregation actions. Calculating a summary measure from a range of ages is a type of aggregation. It works like so:

```{r}
hmd_summary <-
  hmd_completed |> 
  group_by(country, sex, year) |> 
  summarize(e0 = ex[1],
            edag0 = calc_edag(dx,ex),
            .groups = "drop")
```

Visualize it:

```{r}
hmd_summary |> 
  ggplot(aes(x = e0, 
             y = edag0, 
             color = sex)) +
  geom_point(alpha = .1) +
  facet_wrap(~ sex)
```


